---
title: "HW1 Submission"
author: "Mitch Maegaard"
date: "9/12/2018"
output: html_document
---

## How to complete the homework.

If you've made it this far, then you've already downloaded the HW for this week.  I suggest that you keep all of the materials, including this .rmd file,  for the week in one folder.  It will help to set the working directory to the folder that contains the HW materials.  You can do this by opening the rmd file in an RStudio editor window and then using the menu commands Session -> Set Working Directory -> To Source File Location.

You'll be adding R code and typing answers in the designated spaces throughout this document.  At the end of the week you'll submit the .rmd file and the "knitted" Word document to the dropbox on D2L.  



## Problem 1 - Agresti 1.2

### -|-|-|-|-|-|-|-|-|-|-|- Problem 1 Answer -|-|-|-|-|-|-|-|-|-|-|-

a. Nominal  
b. Ordinal  
c. Ordinal  
d. Nominal  
e. Nominal  
f. Ordinal  

---

## Problem 2 - Agresti 1.4

### -|-|-|-|-|-|-|-|-|-|-|- Problem 2 Answer -|-|-|-|-|-|-|-|-|-|-|-

a. The student's number of correct answers on the exam has binomial distribution with mean $100*(.25)=25$ and variance $100(.25)(.75)=18.75$.  
b. Yes, it would be very suprising if the student had 50 correct responses; by guessing, we only expect them to score 25 correctly and the standard deviation is not wide enough (if we constructed a confidence interval) to include a score of 50 within any reasonable limits.

---

## Problem 3 - Agresti 1.6

### -|-|-|-|-|-|-|-|-|-|-|- Problem 3 Answer -|-|-|-|-|-|-|-|-|-|-|-

a. We know $\sum_3n_3=n$, and since we already know $n_1$ and $n_2$, $n_3=3-n_1-n_2$.  
b. When $n=3$, $(3,0,0),(2,1,0),(2,0,1),(1,2,0),(1,0,2),(1,1,1),(0,3,0),(0,0,3),(0,2,1),(0,1,2)$.  
c. $P(n_1,n_2,n_3)=P(1,2,0)=(\frac{3!}{1!2!0!}) 0.25^1 0.50^2 0.25^0 = 0.1875$.  
d. When we look at $n_1$ alone, it has binomial distribution with sample size $n=3$ and $p=\frac{1}{3}$.

```{r}
# calculation for (c)
(factorial(3)/(factorial(1)*factorial(2)*factorial(0)))*(.25)*(.5^2)*(.25^0)
```

---

## Problem 4 - Anti-Anxiety Drugs
A pharmaceutical company is researching a new anti-anxiety drug.  Marketing a new drug is costly, and the company only makes a profit when the drug is effective for more than 80% of patients.  A sample of 60 participants in a drug study showed that 52 responded to the drug. 

a. Find a 95% confidence interval for the population proportion of patients who would respond to the drug.  
b. Do a test at the 5% level to determine if the pharmaceutical company should market the drug.  Include null, alternative, test statistic, p-value, and a practical conclusion in the context of the problem.  
c. Find the test statistics that would be used for both the Wald and Likelihood Ratio Test. (No need to complete other aspects of the test.)

### -|-|-|-|-|-|-|-|-|-|-|- Problem 4 Answer -|-|-|-|-|-|-|-|-|-|-|-

a. Note: we can approximate with normal distribution since $np,npq \geq 5$. A $95\%$ confidence interval can be represented by [0.781, 0.953].  

b.  
$H_0$: The drug is not effective ($\geq 80\%$ of patients) in treating anxiety.  
$H_a$: The new drug is effective in treating anxiety.  
Test stat: $z$ (normal distribution)  
p-value: 0.05  
Conclusion: We are $95\%$ confident that the population proportion of patients who were effectively treated with the new anti-anxiety drug is between 0.781 and 0.953.  

c. Wald: 2.308; Likelihood Ratio Test: 1.837

```{r}
# a.
b <- .8
n <- 60; p <- 52/n
c(n*p, n*p*(1-p))

se <- sqrt((p*(1-p))/n)
alpha <- .95
z <- -qnorm((1-alpha)/2)
low <- p-z*se
high <- p+z*se
c(low, high)

# c.
wald.stat <- ((p-b)/se)^2
l0 <- (factorial(n)/(factorial(52)*factorial(8)))*(b^52)*((1-b)^8)
l1 <- (factorial(n)/(factorial(52)*factorial(8)))*(p^52)*((1-p)^8)
loglik.stat <- -2*log(l0/l1)
c(wald.stat, loglik.stat)
```

---

## Problem 5 - For Graduate Students
In the notes (`02-inference-basics.Rpres`), we mentioned an alternative confidence interval (CI) approach that defines the CI as the set of $\pi_0$ values that would not be rejected when testing. Using the method described, show that with $p=0.04$ and $n=400$, this interval yields $(0.0248, 0.0640)$ (as it did in the notes). *Hint: recognize the quadratic form.*

### -|-|-|-|-|-|-|-|-|-|-|- Problem 5 Answer -|-|-|-|-|-|-|-|-|-|-|-

Original equation: $\frac{p-\pi_0}{\sqrt\frac{\pi_0(1-\pi_0)}{n}} = \pm1.96$  
Rearrange: $p^2-2p\pi_0+\pi_0^2=(\pm1.96)^2\frac{\pi_0(1-\pi_0)}{n}$  
Plug numbers: $p^2 - 2p\pi_0+\pi_0^2=\frac{1.96^2}{n}(\pi_0-\pi_0^2)$  
Rearrange: $p^2-(2p-\frac{1.96^2}{n})\pi_0+(1+\frac{1.96^2}{n})\pi_0^2=0$  
Use quadratic formula to solve for $\pi_0$:

```{r}
p <- 0.04
n <- 400
z <- 1.96

quad.form <- function(a, b, c){
  # input: variables of equation of the form: ax^2+bx=c=0
  # output: lower and upper case values of x
  numerator.base <- sqrt(b^2 - (4*a*c))
  lower <- round((-b - numerator.base) / (2*a), 4)
  upper <- round((-b + numerator.base) / (2*a), 4)
  return(c(lower, upper))
}

a <- 1+(z^2/n)
b <- -((z^2/n)+(2*p))
c <- p^2

quad.form(a,b,c)
```

---
